% Dominic Moylett - December 2015/January 2016
% LaTeX template courtesy of Thomas Mortensson, Ben Collins, Graham Laming and Rob Briggs: https://github.com/thomasmortensson/bristolraspberrypipythontutorial/

\documentclass[twocolumn]{article}
\usepackage[margin=0.5in]{geometry}
\usepackage{listings}
\usepackage{hyperref}
\usepackage[usenames,dvipsnames]{color}
\usepackage{booktabs}
\usepackage{amsmath}

\setlength{\parskip}{1em}

\renewcommand{\familydefault}{\sfdefault}
\definecolor{DarkGreen}{rgb}{0.0,0.4,0.0} % Comment color
\definecolor{highlight}{RGB}{255,251,204} % Code highlight color

\lstdefinestyle{Style1}{ % Define a style for your code snippet, multiple definitions can be made if, for example, you wish to insert multiple code snippets using different programming languages into one document
language=C, % Detects keywords, comments, strings, functions, etc for the language specified
backgroundcolor=\color{highlight}, % Set the background color for the snippet - useful for highlighting
basicstyle=\footnotesize\ttfamily, % The default font size and style of the code
breakatwhitespace=false, % If true, only allows line breaks at white space
breaklines=true, % Automatic line breaking (prevents code from protruding outside the box)
captionpos=b, % Sets the caption position: b for bottom; t for top
commentstyle=\usefont{T1}{pcr}{m}{sl}\color{DarkGreen}, % Style of comments within the code - dark green courier font
deletekeywords={}, % If you want to delete any keywords from the current language separate them by commas
%escapeinside={\%}, % This allows you to escape to LaTeX using the character in the bracket
firstnumber=1, % Line numbers begin at line 1
frame=single, % Frame around the code box, value can be: none, leftline, topline, bottomline, lines, single, shadowbox
frameround=tttt, % Rounds the corners of the frame for the top left, top right, bottom left and bottom right positions
keywordstyle=\color{Blue}\bf, % Functions are bold and blue
morekeywords={VectorSize, SigmaX, Hadamard, CNOT, MeasureBit, Print, measured_value}, % Add any functions no included by default here separated by commas
numbers=left, % Location of line numbers, can take the values of: none, left, right
numbersep=5pt, % Distance of line numbers from the code box
numberstyle=\tiny\color{Gray}, % Style used for line numbers
rulecolor=\color{black}, % Frame border color
showstringspaces=false, % Don't put marks in string spaces
showtabs=false, % Display tabs in the code as lines
stepnumber=1, % The step distance between line numbers, i.e. how often will lines be numbered
stringstyle=\color{Purple}, % Strings are purple
tabsize=2, % Number of spaces per tab in the code
}

\begin{document}
\lstset{style=Style1}

\title{Programming a Quantum Computer} 
\author{Dominic Moylett\\
        	Quantum Engineering Centre for Doctoral Training,\\
		University of Bristol,\\
		\texttt{\href{mailto:dm1905@bristol.ac.uk}{dm1905@bristol.ac.uk}} 
		}
\date{\today} 
\maketitle

\section{What is Quantum Computing?}

Quantum computing was first proposed in the 1980s by a number of physicists, who said that since computers are physical objects, we should build them to take advantage of physical effects. The most bizarre and powerful physical effects have been witnessed in the area of quantum physics, which studies the measurement and interaction of really small particles. In the decades that have followed since then, academics at many institutions -- including the University of Bristol -- have studied how these effects can benefit our technology, and how we can realise this in practice.

In this workshop, we aim to highlight some of the strange effects that occur at the quantum level, and how they can allow our computers to be faster and more powerful than otherwise. To do so, we will simulate a small quantum computer using the Quantum Computing Playground, a website developed by engineers at Google to show off a quantum computer.

All you need to get started is a laptop with Google Chrome installed. Open up Google Chrome and go to the website \url{http://quantumplayground.net/}. Click ``Playground'' in the menu. This will load the playground with a default script. Delete this script as it will not be necessary, and click on the ``2D+Phase'' icon in the top left of the screen. We are now ready to explore the world of quantum computers!

\section{Quantum Bits}

A classical computer is made up of bits: $1$s and $0$s that represent any data the computer stores internally. Quantum computers have a way of storing data as well, known as quantum bits, or {\em qubits}. Because we often work with both classical and quantum bits, we distunguish between them by writing qubits as $|1\rangle$ and $|0\rangle$. These symbols are called {\em kets}, and can be used to describe the whole of quantum mechanics. Note that the text written inside the ket is just a label; $|+\rangle$, $|x\rangle$ and $| \text{cat is alive} \rangle$ are all valid kets too.

In the Quantum Playground, we define our qubits using the command \texttt{VectorSize}, which we write at the top of our code. This specifies the number of qubits we will be using for computation, and the playground initialises these qubits all to the value $|0\rangle$ Try running the code below by typing it into the text box, pressing ``Compile'' and then pressing ``Run'':

\begin{lstlisting}
VectorSize 6
\end{lstlisting}

On the left, this will show us a black image with one white square in the bottom right hand corner. This image describes the state of our qubits. Each square of this image represents a quantum state, and hovering over a square with the mouse will show us the quantum state, and the numbers for Re and Im. The number Re is $0$ for all quantum states except for the state $|0\rangle$, where Re is $1$. We will see what these numbers mean in the next sections.

\section{Measurement}

While images like this showing off our complete quantum state are nice, we cannot look at quantum states this way in practice. Instead, we have to {\em measure} our qubits. We can do this using the \texttt{MeasureBit} command:

\begin{lstlisting}
VectorSize 6
MeasureBit 0
Print "Qubit 0 is |" + measured_value + ">"
\end{lstlisting}

The second line measures the qubit and puts the result in the \texttt{measured\_value} variable. The third line displays the result.

From this we get the output ``Qubit $0$ is $|0\rangle$''. So, if we initialise our quantum state with all qubits set to $|0\rangle$ and we measure one qubit, we get the result that the qubit is  $|0\rangle$.

\section{Quantum Gates}

But data is only one half of computation. We also need be able to modify the data and perform operations on it. Classical computers do this with logic gates, such as the $NOT$ gate, which converts a $NOT(0) = 1$ and $NOT(1) = 0$.

In quantum computing, we also have logic gates, which are called quantum gates. One example of this is the $X$ gate, which is the quantum equivalent of a classical $NOT$ gate: $X|0\rangle = |1\rangle$ and $X|1\rangle = |0\rangle$. Let's try this in the playground:

\begin{lstlisting}
VectorSize 6
SigmaX 0
\end{lstlisting}

Now our image is black with one white square denoting an Re number of 1 for the $|1\rangle$ state. In other words, our qubit has flipped value from $|0\rangle$ to $|1\rangle$. We can also check this properly by measuring our qubit:

\begin{lstlisting}
VectorSize 6
SigmaX 0
MeasureBit 0
Print "Qubit 0 is |" + measured_value + ">"
\end{lstlisting}

Now when we measure our qubit, we find that the result is $|1\rangle$. We can also check the other direction by applying $X$ to our qubit twice before measureing it, and find that $XX|0\rangle = X|1\rangle = |0\rangle$.

\section{Superposition}

Another example of a single qubit gate is the Hadamard gate, $H$. Let's see what this gate does to a qubit in the $|0\rangle$ state:

\begin{lstlisting}
VectorSize 6
Hadamard 0
\end{lstlisting}

This now produces white square for the states $|0\rangle$ and $|1\rangle$, both with Re numbers of $0.707107 \simeq \frac{1}{\sqrt{2}}$. This seems different to classical computing, so let's measure it a few times to see what happens:

\begin{lstlisting}
VectorSize 6
Hadamard 0
MeasureBit 0
Print "Qubit 0 is |" + measured_value + ">"
\end{lstlisting}

Running this multiple times, we see that some times we get $|0\rangle$ and other times we get $|1\rangle$. This is called a superposition, where measuring the qubit could give $|0\rangle$ or $|1\rangle$ purely at random. Pure randomness is already a very useful reason for utilising quantum effects in technology, as classical computers are very deterministic and thus not good at generating random numbers. And this is already in use, a number of online gambling websites use quantum superpositions as one of the ways in which they generate random numbers.

But what determines the probability of us measuring $|0\rangle$ or $|1\rangle$? This is decided by the Re and Im\footnote{For this workshop, we will only focus on the Re numbers and the Im numbers will be $0$.} numbers we have seen. These numbers form the {\em amplitude}s of the quantum state. When we measure a qubit, the probability of measuring a state with amplitude $a$ is $|a|^2$\footnote{The absolute function $|x|$ converts any negative numbers to positive numbers: $|-1| = |1| = 1$}. In our case, the probability of measuring $|0\rangle$ and $|1\rangle$ are both $|\frac{1}{\sqrt{2}}|^2 = \frac{1}{\sqrt{2}}^2 = \frac{1}{2}$. Because the amplitude of each state is $\frac{1}{\sqrt{2}}$, we say that the state is:

$$H|0\rangle = \frac{|0\rangle + |1\rangle}{\sqrt{2}}$$

So we can generate a random bit based on the result of this state. Can we generate more random bits by measuring it multiple times? Let's try:

\begin{lstlisting}
VectorSize 6
Hadamard 0
MeasureBit 0
Print "Qubit 0 is |" + measured_value + ">"
MeasureBit 0
Print "Qubit 0 is |" + measured_value + ">"
\end{lstlisting}

After running this a few times, it seems clear that the first measurement is random, but the second measurement is the same. This is true even if we measure it three, four or a hundred times. This is because once we have measured a qubit in a superposition, that qubit has {\em collapsed} into either $|0\rangle$ or $|1\rangle$, and the randomness has now been lost.

Finally, what happens when we try applying $H$ to a qubit in the $|1\rangle$ state?

\begin{lstlisting}
VectorSize 6
SigmaX 0
Hadamard 0
MeasureBit 0
Print "Qubit 0 is |" + measured_value + ">"
\end{lstlisting}

This seems to produce the same result as before: The qubit is in a superposition, and when we measure it the particle collapses into $|0\rangle$ or $|1\rangle$. So does this produce the same result as $H|0\rangle$? Remove the last two lines of code and look at the image. Now the square for the $|0\rangle$ state is white with an amplitude of $\frac{1}{\sqrt{2}}$, but the $|1\rangle$ state is light blue, with an amplitude of $-\frac{1}{\sqrt{2}}$. Thus this state is:

$$H|1\rangle = \frac{|0\rangle - |1\rangle}{\sqrt{2}}$$

This explains why the results when measuring look so similar, as the probability of getting $|0\rangle$ or $|1\rangle$ is still $\frac{1}{2}$.

You might wonder at this point what other quantum state we can possibly make with a single qubit. The answer is that we can make any qubit of the form $a|0\rangle + b|1\rangle$, where $|a|^2 + |b|^2 = 1$.

\section{Multiple Qubits}

Now that we have seen single qubits, let's look at multiple qubits. Multiple qubits act similarly to multiple classical bits, where we can apply quantum gates to the individual qubits. We represent these qubits in the {\em tensor product}. For example, suppose the first qubit is $|0\rangle$ and the second is $|1\rangle$. We would write this state as $|0\rangle \otimes |1\rangle$ and often shorten it to $|01\rangle$.

So for example, we can apply an $X$ gate to the first and second qubits as follows:

\begin{lstlisting}
VectorSize 6
SigmaX 0
SigmaX 1
MeasureBit 0
Print "Qubit 0 is |" + measured_value + ">"
MeasureBit 1
Print "Qubit 1 is |" + measured_value + ">"
\end{lstlisting}

We find that they are in the $|11\rangle$ similarly to when we were acting only on individual qubits. Likewise, if we applied nothing to them and measured qubits $0$ and $1$ we would find that they were both in the $|00\rangle$ state.

And similarly again, if we apply a Hadamard gate to both qubits, we would find that they were in their own superpositions and would produce random measurement results, independently of each other:

\begin{lstlisting}
VectorSize 6
Hadamard 0
Hadamard 1
MeasureBit 0
Print "Qubit 0 is |" + measured_value + ">"
MeasureBit 1
Print "Qubit 1 is |" + measured_value + ">"
\end{lstlisting}

Like with single qubits, this means we can produce any superposition over two qubits. Thus our state can be any form of $a|00\rangle + b|01\rangle + c|10\rangle + d|11\rangle$ where $|a|^2 + |b|^2 + |c|^2 + |d|^2 = 1$

\section{Entanglement}

Not all classical gates are single bit. The classical AND gate for example, takes two classical bits as input and produces one classical bit as output.

We also have multiple qubit quantum gates. One of particular importance is the Controlled-NOT gate, or $CNOT$, which takes a control qubit as the first argument and a target qubit as the second. Let's see what happens when the control qubit is $|0\rangle$:

\begin{lstlisting}
VectorSize 6
CNOT 0, 1
MeasureBit 0
Print "Qubit 0 is |" + measured_value + ">"
MeasureBit 1
Print "Qubit 1 is |" + measured_value + ">"
\end{lstlisting}

It seems that $CNOT|00\rangle = |00\rangle$. If we were to apply an $X$ gate to qubit $1$ before the $CNOT$ we would find that $CNOT|01\rangle = |01\rangle$. So when the control qubit is $|0\rangle$, the $CNOT$ gate does nothing. What if the control qubit is $|1\rangle$?

\begin{lstlisting}
VectorSize 6
SigmaX 0
CNOT 0, 1
MeasureBit 0
Print "Qubit 0 is |" + measured_value + ">"
MeasureBit 1
Print "Qubit 1 is |" + measured_value + ">"
\end{lstlisting}

Now our target qubit has been flipped, so we can see that $CNOT|10\rangle = |11\rangle$. And again, if we were to apply an $X$ gate to our target qubit too, we would find that $CNOT|11\rangle = |10\rangle$.

But these aren't the only possible states our control qubit can take. What if our control qubit was in a superposition?

\begin{lstlisting}
VectorSize 6
Hadamard 0
CNOT 0, 1
MeasureBit 0
Print "Qubit 0 is |" + measured_value + ">"
MeasureBit 1
Print "Qubit 1 is |" + measured_value + ">"
\end{lstlisting}

This produces a very different result. It now appears that our measurement of qubit $0$ is random, but our measurement of qubit $1$ is always the same state as qubit $0$. If we measure qubit $1$ first; then our measurement of qubit $1$ will be random, but our measurement of qubit $0$ will always be the same state as qubit $1$. It seems that our qubits are in a superposition together, and if one collapses then the other collapses with it.

This effect is called {\em entanglement}, and is one of the strangest results of quantum mechanics. When Einstein, Podolsky and Rosen first discovered the theory of entangled states, Einstein was skeptical of this being possible. And he had good reason to be, as entangled states seem to collapse at exactly the same time, even if one qubit was at the At-Bristol Science Centre and the other was at the Digimakers Moon Base\footnote{Construction pending}. Classically this simultaneous collapse should be impossible, as that would require one qubit to send a message to the other faster than the speed of light, violating Einstein's laws of relativity.

While we haven't been able to perform the entanglement of qubits between the At-Bristol Science Centre and the Digimakers Moon Base, physicists at Delft University of Technology in The Netherlands have managed to successfully entangle qubits and witness these effects over a kilometre apart from each other. So how does entanglement work if it cannot be done classically? The answer is unknown, and there are many ideas based on different interpretations of quantum mechanics.

Does this mean that entanglement allows us to send messages to each other instantly then? While it would be nice if this was the case, the answer is no. To see this, suppose we entangled two qubits, one here with you and the other with someone else on the Digimakers Moon Base. The person on the moon might try to send a message to you by collapsing their qubit, thus collapsing your's at the same time. Unfortunately, you cannot see whether not they collapsed their qubit without looking at, and thus collapsing, your own qubit. So when you measure your qubit, it is impossible to tell whether it collapsed because of your measurement, or if they had already measured their own. So while entanglement itself seems to violate the laws of relativity, we cannot use it to violate these laws ourselves.

\section{Deutsch's Algorithm}

We are now ready to program our first quantum algorithm and see one problem where quantum computers can truly outperform classical ones. This problem was first devised by David Deutsch, and is stated as follows: Suppose we have a function $f$, which takes one classical bit as input and returns one classical bit as output. In other words, $f(0)$ is either $0$ or $1$ and $f(1)$ is either $0$ or $1$. Deutsch's problem is to determine if $f(0) = f(1)$.

On a classical computer, we would need at least two queries to determine this with certainty: One to figure out what $f(0)$ is and one to figure out what $f(1)$ is. But can we do better with a quantum version of this function?

Deutsch's Algorithm can be implemented as follows:

\begin{lstlisting}
VectorSize 6
SigmaX 1
Hadamard 0
Hadamard 1
//Insert quantum version of function here
Hadamard 0
MeasureBit 0
Print "Qubit 0 is |" + measured_value + ">"
\end{lstlisting}

For line 5, we pick what operation to run based on what we want $f(0)$ and $f(1)$ to be. This quantum operator is called $U_f$, and is formally stated as $U_f|x\rangle|y\rangle = |x\rangle|y \oplus f(x)\rangle$, where $\oplus$ is the exclusive-OR (XOR) gate. It might seem difficult to implement a quantum gate like this, but we can actually do it very easily, based on what the function $f$ is.

Case $1$: If you want $f(0) = 0 \text{ and } f(1) = 0$, then delete line $5$.

Case $2$: If you want $f(0) = 0 \text{ and } f(1) = 1$, then replace line $5$ with:

\begin{lstlisting}
CNOT 0, 1
\end{lstlisting}

Case $3$: If you want $f(0) = 1 \text{ and } f(1) = 0$, then replace line $5$ with:

\begin{lstlisting}
CNOT 0, 1
SigmaX 1
\end{lstlisting}

Case $4$: Or if you want $f(0) = 1 \text{ and } f(1) = 0$, then replace line $5$ with:

\begin{lstlisting}
SigmaX 1
\end{lstlisting}

If we run the full program, we find that if $f(0) = f(1)$ (cases $1$ and $4$) then qubit $0$ when measured is in the $|0\rangle$ state, and if $f(0) \neq f(1)$ (cases $1$ and $2$) then qubit $0$ when measured is in the $|1\rangle$ state. Thus we have managed to figure out if $f(0) = f(1)$ with only one query of our function. Later, David Deutsch and Richard Josza showed that a more general problem would require an exponential number of queries on a classical computer, but just one query on a quantum computer.

You might wonder what the point of this was if we can already see the code, as in this case. In this case, we can see the code that makes up $U_f$, and can thus figure out if $f(0) = f(1)$ based on just reading that. But we might not always have access to the code; we might just be given a black box which we know implements $U_f|x\rangle|y\rangle = |x\rangle|y \oplus f(x)\rangle$, without knowing what exactly $f$ is. Deutsch's Algorithm works in that case as well.

This problem doesn't sound very useful, but there are many more practical problems for which quantum computers can do better than classical computers:

\begin{itemize}
\item Charles Bennett and Stephen Wiesner showed that you can send two classical bits from one person to another by just sending one qubit from one to the other.
\item Searching through an unsorted list of items $n$ for a particular item requires at most $n$ classical queries -- one for each item. Lov Grover showed that on a quantum computer, we can do this with $\sqrt{n}$ queries.
\item Peter Shor showed that you can efficiently find the prime factors of an integer on a quantum computer. This problem is believed to be hard to solve on a classical computer, because if it was easy then the RSA encryption scheme would be broken.
\end{itemize}

\section{Wait, where's the Physics?}

We have seen the theory behind a lot of quantum effects today, but not much on where these quantum effects actually occurr. When do superpositions or entanglement happen in the real world? I have avoided talking about real life experiments partially because they become more complex as implementation details are added, but also because quantum effects happen in many different cases.

When the first quantum effects were witnessed, they were seen in experiments involving light. This could be explained by concluding that light was a wave. But we then saw the same effects with electrons, which are definitely not waves. And we have seen them with larger and larger particles ever since. The largest particles we have seen quantum effects on are bucky-balls: Molecules made up of sixty carbon atoms joined together. And for all of these particles, the quantum effects we have witnessed with them can be explained using theory similar to what we have seen today.

\section{When will I be able to Buy a Quantum Computer?}

Depending on what you mean by a quantum computer, you can buy one right now! The Canadian company D-Wave Systems already sell quantum computers, but these computers are only built for solving a very specific problem called quantum annealing. But they are on the market, and Google are currently seeing how useful the quantum annealer they bought from D-Wave is for their business.

But what is more interesting in the Physics community is when will we see a {\em universal quantum computer} -- a computer that can perform any quantum computation -- on the market. And the answer is we don't know, but you can use one right now! The Centre for Quantum Photonics at the University of Bristol connected their two qubit universal quantum computer to the Internet, in what they call the Quantum in the Cloud project.

But two qubits is not very useful either: When are we going to get many qubit universal quantum computers? And the answer is we don't know, but potentially soon. John Martinis of the University of California, Santa Barbara has partnered up with Google and is hoping to develop a $100$ qubit universal quantum computer within the next 2-3 years. Microsoft Research predicts a quantum computer to be built within the next ten years. Others have predicted timescales ranging from five years to forty years. Regardless of when it comes, one thing is certain: There's a lot of excitement surrounding quantum computers right here and now.

\section{Further Reading}

I first heard about quantum computing at an open day at the University of Bristol. At the open day I was given a booklet written by Ashley Montanaro of the Quantum Computation and Information Group at the University of Bristol, which explains many of the concepts discussed here -- including Deutsch's Algorithm -- from an even more theoretical perspective. That booklet is available here: \url{http://www.cs.bris.ac.uk/~montanar/gameshow.pdf}

Ashley also gave a guest lecture on quantum computing from both a theoretical and practical perspective, the slides of which are available here:\url{http://www.maths.bris.ac.uk/~csxam/teaching/history.pdf}

Juan Miguel Arrazola of the Institute for Quatum Computing at the University of Waterloo has been writing a collection of blog posts explaining quantum mechanics to everyone. Part one is here: \url{https://uwaterloo.ca/institute-for-quantum-computing/blog/post/anyone-can-understand-quantum-mechanics-part-1}; part two is here: \url{https://uwaterloo.ca/institute-for-quantum-computing/blog/post/anyone-can-understand-quantum-mechanics-part-2}; and part three will be coming soon, so make sure to keep an eye on their blog.

Note that while these are the easier aspects of quantum mechanics to understand, a lot of quantum mechanics in general rely on topics that are taught at GCSE or A Level Mathematics, so be cautious about running out and buying a copy of Nielsen and Chuang's Quantum Computation and Quantum Information just yet! But if you want to try and get ahead, some very useful mathematical concepts to learn include:

\begin{itemize}
\item Trigonometry ($\sin, \cos$, trigonometric identities...)
\item Complex numbers ($e^{i\theta} = \cos\theta + i\sin\theta$, complex conjugates...)
\item Linear algebra (Matrices, vectors, eigenvalues \& eigenvectors...)
\end{itemize}

All of these concepts come up on a regular basis in quantum computing, so if you want more of a reason to study these concepts, now's your chance.

\end{document} 